Handling Recursion in Generic Programming Using Closed Type Families

### 2018-11-16 lncs-paper.tex
More examples of generic functions, address some small-scale remarks from the TFP reviews

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{url}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{fancybox}
\usepackage{xcolor}

\lstdefinestyle{fancy}{
language=Haskell,
deletekeywords={Int,Bool,String,Functor,Char,True,False,
  Maybe,Nothing,Just,Show,show,Monad,catMaybes,const,id},
morekeywords={family},
basicstyle=\color{cbasic}\rmfamily,
identifierstyle=\color{cid}\rmfamily\itshape,
keywordstyle=\color{ckey}\rmfamily\bfseries,
xleftmargin=\parindent,
aboveskip=0.2cm,
belowskip=0.2cm,
basewidth=0.48em,
numberbychapter=false,
showstringspaces=false,
moredelim=[is][\color{ccom}\itshape]{/}{/},
escapeinside={(/}{/)},
literate={:: }{{$::$}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\scriptsize\bf\dots}}2
{>>>}{{\scriptsize\comp}}3 {>=>}{{\scriptsize\kleisli}}3
{.1}{{$_\mathit{1}$}}1 {.2}{{$_\mathit{2}$}}1
{.n}{{$_\mathit{n}$}}1 {.k}{{$_\mathit{k}$}}1
{.Aux}{{\large$_\mathit{\hspace{0,05cm} Aux}$}}3
{..Aux}{{\large$_\mathit{\hspace{-0,03cm} Aux}$}}3
{@}{{@\hspace{-0,045cm}}}2
{_}{{\underline{\hspace{0.45em}}}}1
{+}{{\scriptsize+}}1
{.-}{{\nhs-}}1
{-.}{{-\nhs}}1
}

@@

\newcommand{\hs}{\hspace{0.1cm}}
\newcommand{\nhs}{\hspace{-0.06cm}}

\newcommand{\tg}{\textgreater}
\newcommand{\Ra}{$\Rightarrow$}
\newcommand{\ra}{$\rightarrow$}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}}
\newcommand{\comp}{\tg\nhs\tg\nhs\tg}
\newcommand{\kleisli}{\tg\nhs=\nhs\tg}
\newcommand{\ann}{:\texttt{\nhs}:}
\newcommand{\ds}{.\nhs.\nhs.}

@@

\begin{document}
%
\title{Handling Recursion in Generic Programming Using Closed Type Families}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Anna Bolotina\inst{1} \and
Artem Pelenitsyn\inst{2}
}
%
%\authorrunning{F. Author and S. Author}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Southern Federal University, Rostov-on-Don, Russia\\
\email{bolotina@sfedu.ru} \and
Czech Technical University in Prague, Prague, Czech Republic\\
\email{pelenart@fit.cvut.cz}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express their recursive structure. However, some other approaches, especially the ones based on sums of products, avoid the fixed point encoding. They facilitate implementation of generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues otherwise. A widely used and unwelcome solution to the problem resorts to overlapping instances. In this paper, we present an alternative approach that uses closed type families to eliminate the need of overlap for handling recursion in datatypes. Moreover, we show that this idiom allows for generic programming with families of mutually recursive datatypes.

@@

There are many generic functions that deal with the recursive knots when traversing the structure of datatypes. Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}; more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}. For handling recursion, several generic programming approaches express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}. The SOP view naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise becomes unhandy.

@@

The problem of using overlapping instances was first addressed by Kiselyov~et~al.~\cite{Kiselyov2004}. Their technique for avoiding overlap relies on a Haskell 98 extension for functional dependencies. The solution proposes two variants of defining a type-level equality predicate, a type class, and then systematically localizes overlap by circumventing it with that predicate. The first version of type equality maps types to unique type representations and compares them. That variant and its later implementation with type families~\cite{Kiselyov2012} fully eliminate \textsf{OverlappingInstances}. Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC. The most generic solution for type equality, the second version, again makes use of overlapping instances, however.

@@

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to represent a datatype as an isomorphic code whose kind is a list of lists of types. The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
\begin{lstlisting}[style=fancy][style=fancy]
type family Code (a :: *) :: [[*]]
\end{lstlisting}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.

@@

The definitions of \K{NS} and \K{NP} are kind polymorphic. The index list is allowed to contain types of arbitrary kind \K{k}, since \K{k} turns to \K{*} by applying the type constructor \K{f}.
Basic instantiations of type parameter \K{f} found in SOP are identity functor \K{I}, that is, a type-level equivalent for \K{id} function, and a constant functor \K{K}, an analogue of \K{const}:
\begin{lstlisting}[style=fancy]
newtype I (a :: *)          = I {unI :: a}
newtype K (a :: *) (b :: k) = K {unK :: a}
\end{lstlisting}
If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[a]}. Here is an example value of type \K{NP I}:
\begin{lstlisting}[style=fancy]
I 5 :* I True :* I 'x' :* Nil :: NP I '[Int, Bool, Char]
\end{lstlisting}

We turn to the sum definition now. The constructor \K{S} of \K{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \K{Z} stores the payload of type \K{f x}. For example, the following chooses the third element of a sum:
\begin{lstlisting}[style=fancy]
S (S (Z (I 5))) :: NS I '[Char, Bool, Int, Bool]
\end{lstlisting}
The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.

With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to} witnessing the isomorphism between a datatype and its generic representation:
\begin{lstlisting}[style=fancy]
type Rep a = SOP I (Code a)

@@

We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition as irrelevant to our work. Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following. Therefore, it is worth noting that \K{All} applies a particular constraint (e.g. \K{SListI} above) to each member of a list of types. The usage of constraints as type arguments is allowed due to the \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.

@@

The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending the current element to the result list if its type is the same as of the original term, otherwise skipping the element. We use GHC's \textsf{TypeAppications} extension to pass that type.

@@

Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type. There is no clear path to this at the moment. Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
\begin{lstlisting}[style=fancy]
class Subterms a (xs :: [*]) where
  subtermsNP :: NP I xs -> [a]

@@

Although the approach works, as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap. This would remove the complexity overhead introduced by the approach, as we have mentioned in the introduction.

@@

Below we show two variants of completing the \K{subterms} implementation. The low-level version performs pattern matching on the structure of \K{NS} and \K{NP}, as we do in Section~\ref{subsec:recursion-problem}. For this variant, we reimplement \K{subtermsNP}. The different implementation employes powerful machinery of high-level combinators---one of the major advantages provided by \textsf{generics-sop}. Although, those high-level functions become one extra layer of complexity.

@@

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for the task of \K{subterms}. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

\subsection{Generic show}
\label{subsec:gshow}

@@

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \K{Show}, and invoke its \K{show} function. Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \K{gshow}. Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
\begin{lstlisting}[style=fancy]
class CaseShow (eq :: Bool) (a :: *) (b :: *) where
  caseShow' :: b -> String

@@

The resulting function \K{gshow} is a subject of a number of constraints abbreviated by a \K{GShow} synonym:
\begin{lstlisting}[style=fancy]
type GShow a = (Generic a, HasDatatypeInfo a,
                All2 (CaseRecShow a) (Code a))

@@

We define \K{gshow} mutually recursive with \K{caseShow}. The full implemenentation of the function \K{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.

@@

"Node (Leaf True) (Leaf False)"
\end{lstlisting}
Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}:
\begin{lstlisting}[style=fancy]
instance Show T where
  show = gshow
\end{lstlisting}
This is a consequence of \K{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \K{Show} constraint for all knots in the datatype structure.

@@

This machinery employes the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec..Aux}. The predicate \K{P} generalizes \K{Equal} to an arbitrary relation on types, so that one can model, e.g., subtyping for some domain-specific language. The introduction of the associated type class \K{DispatchRec} provides a handy interface for dispatch. The dispatch function \K{dispatchRec} decides on one of the functions \K{f.1} and \K{f.2}---depending on whether the predicate \K{P} is true for types \K{a.1}, \dots, \K{a.k}, \K{b}.

For consumer functions, such as \K{subterms} and \K{gshow}, the type \K{X b} typically is instantiated with \K{b -> T} for some datatype \K{T}. The pattern fits equally well for defining producers, such as \K{garbitrary} from the \textsf{basic-sop} package, where \K{X b} would be instantiated with \K{Gen b}.

@@

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{url}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{listings}
\usepackage{fancybox}
\usepackage{xcolor}

\lstdefinestyle{fancy}{
language=Haskell,
deletekeywords={Int,Bool,String,Functor,Char,True,False,
  Maybe,Nothing,Just,Show,show,Monad,catMaybes,const,id},
morekeywords={family},
basicstyle=\color{cbasic}\rmfamily,
identifierstyle=\color{cid}\rmfamily\itshape,
keywordstyle=\color{ckey}\rmfamily\bfseries,
xleftmargin=\parindent,
aboveskip=0.2cm,
belowskip=0.2cm,
basewidth=0.48em,
numberbychapter=false,
showstringspaces=false,
moredelim=[is][\color{ccom}\itshape]{/}{/},
escapeinside={(/}{/)},
literate={:: }{{$::$}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\scriptsize\bf\dots}}2
{>>>}{{\footnotesize\comp}}3 {>=>}{{\footnotesize\kleisli}}3
{.1}{{$_\mathit{1}$}}1 {.2}{{$_\mathit{2}$}}1
{.3}{{$_\mathit{3}$}}1
{.i}{{$_\mathit{i}$}}1 {.j}{{$_\mathit{j}$}}1
{.n}{{$_\mathit{n}$}}1 {.k}{{$_\mathit{k}$}}1
{.R}{{\scriptsize$_\mathit{\hspace{0,07cm} \boldsymbol{R}}$}}1
{..R}{{\scriptsize$_\mathit{\hspace{-0,02cm} \boldsymbol{R}}$}}1
{,R}{{\scriptsize$_\mathit{\hspace{-0,05cm} \boldsymbol{R}}$}}1
{,,R}{{\scriptsize$_\mathit{\boldsymbol{R}}$}}1
{.Aux}{{\large$_\mathit{\hspace{0,05cm} Aux}$}}3
{.AExpr}{{\large$_\mathit{\hspace{0,05cm} AExpr}$}}5
{..Aux}{{\large$_\mathit{\hspace{-0,03cm} Aux}$}}3
{@}{{@\hspace{-0,045cm}}}2
{_}{{\underline{\hspace{0.45em}}}}1
{[.}{{[\hspace{-0,06cm}}}1
{[..}{{[\hspace{-0,1cm}}}1
{,,]}{{\hspace{-0,1cm}}]}1
{,,,]}{{\hspace{-0,16cm}}]}1
{,]}{{\hspace{-0,06cm}]}}1
{+}{{\scriptsize+}}1
{,*.}{{\nhs$*$\nhs}}1
{+.}{{\footnotesize+}}1
{.+.}{{\nhs\footnotesize+\nhs}}1
{*.}{{\footnotesize$\times$}}1
{.*.}{{\nhs\footnotesize$\times$\nhs}}1
{.-}{{\nhs-}}1
{-.}{{-\nhs}}1
}

@@

\newcommand{\hs}{\hspace{0.1cm}}
\newcommand{\nhs}{\hspace{-0.06cm}}
\newcommand{\Nhs}{\hspace{-0.1cm}}

\newcommand{\tg}{\textgreater}
\newcommand{\Ra}{$\Rightarrow$}
\newcommand{\ra}{$\rightarrow$}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}}
\newcommand{\comp}{\tg\Nhs\tg\Nhs\tg}
\newcommand{\kleisli}{\tg\Nhs=\Nhs\tg}
\newcommand{\ann}{:\texttt{\nhs}:}
\newcommand{\ds}{.\nhs.\nhs.}

@@

\begin{document}
%
\title{Handling Recursion in Generic Programming Using Closed Type Families}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Anna Bolotina\inst{1} \and
Artem Pelenitsyn\inst{2}
}
%
%\authorrunning{F. Author and S. Author}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Southern Federal University, Rostov-on-Don, Russia\\
\email{bolotina@sfedu.ru} \and
Northeastern University, Boston, USA\\
\email{artem@ccs.neu.edu}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express their recursive structure. However, some other approaches, especially the ones based on sums of products, avoid the fixed point encoding. They facilitate implementation of generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues otherwise. A widely used and unwelcome solution to the problem resorts to overlapping instances. In this paper, we present an alternative approach that uses closed type families to eliminate the need of overlap for handling recursion in datatypes. Moreover, we show that this idiom allows for generic programming with families of mutually recursive datatypes.

@@

There are many generic functions that deal with the recursive knots when traversing the structure of datatypes. Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}; a more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}. For handling recursion, several generic programming approaches express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}. The SOP view naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise becomes unhandy.

@@

The problem of using overlapping instances was first addressed by Kiselyov~et~al.~\cite{Kiselyov2004}. Their technique for avoiding overlap relies on a Haskell 98 extension for functional dependencies. The solution proposes two variants of defining a type-level equality predicate, a type class, and then systematically localizes overlap by circumventing it with that predicate. The first version of type equality maps types to unique type representations and compares them. That variant and its later implementation with type families~\cite{Kiselyov2012} fully eliminate \textsf{OverlappingInstances}. Despite this, each type needs a representation instance to be derived---by means of Template Haskell or GHC. The most generic solution for type equality, the second version, again makes use of overlapping instances, however.

@@

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to represent a datatype as an isomorphic code whose kind is a list of lists of types. The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
\begin{lstlisting}[style=fancy][style=fancy]
type family Code (a :: *) :: [[*]]
\end{lstlisting}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list represents an $n$-ary product, isomorphic to a sequence of constructor arguments, while the outer list, representing an $n$-ary sum, corresponds to a choice of a particular constructor.

@@

The definitions of \K{NS} and \K{NP} are kind polymorphic. The index list is allowed to contain types of arbitrary kind \K{k}, since \K{k} turns to \K{*} by applying the type constructor \K{f}.
Basic instantiations of type parameter \K{f} found in SOP are identity functor \K{I}, that is, a type-level equivalent for \K{id} function, and a constant functor \K{K}, an analogue of \K{const}:
\begin{lstlisting}[style=fancy]
newtype I (a :: *)          = I {unI :: a}
newtype K (a :: *) (b :: k) = K {unK :: a}
\end{lstlisting}
If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[.a]}. Here is an example value of type \K{NP I}:
\begin{lstlisting}[style=fancy]
I 5 :* I True :* I 'x' :* Nil :: NP I '[Int, Bool, Char]
\end{lstlisting}

We turn to the sum definition now. The constructors \K{S} and \K{Z} of \K{NS} allow to represent choices from a sum as indices in the form of Peano numbers. Firstly, the application of \K{Z} represents the choice of the first component of the sum. Secondly, \K{S} is the successor constructor. That is, \K{S . Z} chooses the second component, and so on up to the length of the index list \K{xs}. Given a sum of type \K{NS f xs}, the $i$-th choice has type \K{f x}, where \K{x} is the $i$-th element in \K{xs}. For example, the following represents the third choice from a sum:
\begin{lstlisting}[style=fancy]
S (S (Z (I 5))) :: NS I '[Char, Bool, Int, Bool]
\end{lstlisting}
Putting this together, each term of a particular datatype can be represented as the choice from the sum of products, representing that datatype. The choice matches the index of its particular constructor in the index list, and stores the product, representing arguments of that constructor.

With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to}, witnessing the isomorphism between a datatype and its generic representation:
\begin{lstlisting}[style=fancy]
type Rep a = SOP I (Code a)

@@

We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition, as it is irrelevant to our work. Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following. Therefore, it is worth noting that \K{All} applies a particular constraint (e.g. \K{SListI} above) to each member of a list of types. The usage of constraints as type arguments is allowed due to the \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.

@@

The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending the current element to the result list if its type is the same as of the original term, otherwise skipping the element. We use GHC's \textsf{TypeApplications} extension to pass that type.

@@

Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type. There is no clear path to this at the moment. Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermsNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
\begin{lstlisting}[style=fancy]
class Subterms a (xs :: [*]) where
  subtermsNP :: NP I xs -> [a]

@@

Although the approach works, as exemplified by a number of the packages on Hackage, we aim to release generic programs from overlap. This would remove the complexity overhead introduced by the approach, as we have mentioned in the introduction.

@@

Below we show two variants of completing the \K{subterms} implementation. The low-level version performs pattern matching on the structure of \K{NS} and \K{NP}, as we do in Section~\ref{subsec:recursion-problem}. For this variant, we reimplement \K{subtermsNP}. The different implementation employs powerful machinery of high-level combinators---one of the major advantages provided by \textsf{generics-sop}. Although, those high-level functions become one extra layer of complexity.

@@

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for the task of \K{subterms}. We give other examples showing how to adapt our idiom to different scenarios in the following subsections.

\subsection{Generic \K{show}}
\label{subsec:gshow}

@@

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \K{Show}, and invoke its \K{show} function. Whereas in the case of the recursive position, we apply our generic function \K{gshow}. Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
\begin{lstlisting}[style=fancy]
class CaseShow (eq :: Bool) (a :: *) (b :: *) where
  caseShow' :: b -> String

@@

The resulting function \K{gshow} is subject to a number of constraints abbreviated by a \K{GShow} synonym:
\begin{lstlisting}[style=fancy]
type GShow a = (Generic a, HasDatatypeInfo a,
                All2 (CaseRecShow a) (Code a))

@@

We define \K{gshow} mutually recursive with \K{caseShow}. The full implementation of the function \K{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.

@@

"Node (Leaf True) (Leaf False)"
\end{lstlisting}

Our implementation of \K{gshow} obviates one drawback of its analogue from \textsf{basic-sop}. This drawback is that \K{gshow} from \textsf{basic-sop} does not treat recursive positions separately, and therefore requires the \K{Show} constraint for all knots in the datatype structure:
\begin{lstlisting}[style=fancy]
gshow :: ∀a. (Generic a, HasDatatypeInfo a, All2 Show (Code a))
      => a -> String
\end{lstlisting}
As a consequence, \textsf{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}, requiring an additional \K{Show} instance declaration for \K{T}:
\begin{lstlisting}[style=fancy]
instance Show T where
  show = gshow
\end{lstlisting}
As we have shown,  the variant of the function \K{gshow}, distinguishing the recursion cases, can be used directly, without this extra declaration. This takes advantage of employing the polytypic instance of \K{CaseRecShow} for all datatypes.

\subsection{Generic recursion schemes}
\label{subsec:rec-schemes}

Recursion schemes, such as \K{fold}~\cite{Meijer1991,MuRec2009} and \K{compos}~\cite{Bringert2006}, are classical examples of generic functions that are usually treated by a fixed point view. In this subsection, we show how they can be defined in the SOP view, adopting our approach.

\subsubsection{Generic \K{compos}}

The function \K{compos} is a traversal combinator for defining compositional functions. It applies a given function to all immediate children of a given term ot type \K{T}:
\begin{lstlisting}[style=fancy]
compos :: (T -> T) -> T -> T
\end{lstlisting}
This can be used, for example, with a datatype \K{Expr} for defining the \K{renameVar} function, updating all variables in a given expression without affecting its other subexpressions:
\begin{lstlisting}[style=fancy]
data Expr = EAbs String Expr
          | EApp Expr Expr
          | EVar String

renameVar :: Expr -> Expr
renameVar (EVar x) = EVar $ x ++ "_"
renameVar e        = compos renameVar e
\end{lstlisting}

The implementation of the generic function \K{gcompos} employs a class for resolving recursion cases, which now looks as the following:
\begin{lstlisting}[style=fancy]
class CaseCompos.Aux (eq :: Bool) (a :: *) (x :: *) (y :: *) where
  caseCompos.Aux :: (a -> a) -> I x -> I y

instance CaseCompos.Aux 'False a x x where
  caseCompos.Aux _ = id
instance CaseCompos.Aux 'True  a a a where
  caseCompos.Aux f = I . f . unI
\end{lstlisting}
The function \K{caseCompos}, an abbreviation for \K{caseCompos..Aux @(Equal a x)}, is supposed to be applied to each component of the product, representing arguments of a chosen constructor. Given a component of type \K{I x}, it results in the component of different type \K{I y}. (The implementation below makes use of a higher-order function, whose type does not admit that \K{x} and \K{y} can be unified here.) The first case in this definition corresponds to mapping over a constant, which should not be changed. The second case is the application of the given function \K{f} to a recursive position.

The rest of the work is handled by \textsf{generics-sop}'s function \K{trans_SOP}:
\begin{lstlisting}[style=fancy]
trans_SOP :: AllZip2 c xss yss
          => proxy c -> (forall x y. c x y => f x -> g y)
          -> SOP f xss -> SOP g yss
\end{lstlisting}
This function transforms the generic representation of a term, applying a given function to every component of a chosen product. \K{AllZip2} here is analogous to \K{All2}, zipping two lists of lists with the binary constraint \K{c}.

Finally, the definition of \K{gcompos}, using \K{trans_SOP} with \K{caseCompos}, is following, with \K{CaseCompos} abbreviating the use of \K{CaseCompos..Aux}:
\begin{lstlisting}[style=fancy]
type GCompos a = (Generic a, AllZip2 (CaseCompos a) (Code a) (Code a))

gcompos :: ∀a. GCompos a => (a -> a) -> a -> a
gcompos f = to . composSOP . from
  where
    composSOP = trans_SOP (Proxy @(CaseCompos a)) $ caseCompos f
\end{lstlisting}

\subsubsection{Generic \K{fold}}

A fold is a higher-order function that recursively deconstructs a term, using a combining operation on its structure to aggregate the result. For example, consider a fold for a datatype of arithmetic expressions \K{AExpr}:
\begin{lstlisting}[style=fancy]
data AExpr = EConst Int
           | EAdd AExpr AExpr
           | EMul AExpr AExpr

fold.AExpr :: (Int -> r, r -> r -> r, r -> r -> r) -> AExpr -> r
fold.AExpr f@(f.1, f.2, f.3) e = case e of
  EConst x -> f.1 x
  EAdd l r -> f.2 (fold.AExpr f l) (fold.AExpr f r)
  EMul l r -> f.3 (fold.AExpr f l) (fold.AExpr f r)
\end{lstlisting}
Given \K{f}---called an \emph{algebra}---it proceeds through the recursive structure of a given term \K{e}, applying its components \K{f.1}, \K{f.2}, and \K{f.3} to elements of \K{e} and combining the results. This can be used to evaluate expressions. Consider:
\begin{lstlisting}[style=fancy]
example = EAdd (EMul (EConst 3) (EConst 2))
               (EAdd (EMul (EConst 2) (EConst 2))
                     (EConst 5))
\end{lstlisting}
The call
\begin{lstlisting}[style=fancy]
fold.AExpr (id, (+), (*)) example
\end{lstlisting}
yields 15.

In the example above, \K{(Int -> r, r -> r -> r, r -> r -> r)} is a type of algebras for the datatype \K{AExpr} to a result type \K{r}. This type represents the underlying recursive structure of \K{AExpr}. In particular, the constructors of \K{AExpr} form an algebra for expressions, which is isomorphic to \K{AExpr}:
\begin{lstlisting}[style=fancy]
(EConst, EAdd, EMul) :: (Int -> AExpr,
                         AExpr -> AExpr -> AExpr,
                         AExpr -> AExpr -> AExpr)
\end{lstlisting}
A type of algebras for a particular datatype \K{a} can be computed generically from its representation code. The result type \K{r} should occur in this type whenever \K{a} (a recursive position) occurs in the code. In the following, the type families \K{AlgebraS} and \K{AlgebraP} are employed to compute a type of algebras for a given datatype \K{a} and a result type \K{r}.
\begin{lstlisting}[style=fancy]
type family AlgebraS (code :: [[*]]) a r :: [*] where
  AlgebraS (xs ': xss) a r = (AlgebraP xs a r ': AlgebraS xss a r)
  AlgebraS '[]         _ _ = '[]

type family AlgebraP (xs :: [*]) a r where
  AlgebraP (a ': xs) a r = r -> AlgebraP xs a r
  AlgebraP (x ': xs) a r = x -> AlgebraP xs a r
  AlgebraP '[]       _ r = r

type Algebra a r = NP I (AlgebraS (Code a) a r)
\end{lstlisting}
The calculated type of algebras \K{Algebra a r} is meant to be an \K{NP} of its components, which are constructed by \K{AlgebraP} from each product in the code. All the components, being applied to the elements of a datatype structure, return a result of type \K{r}, as handled by the last case in the \K{AlgebraP} definition.

We turn to defining a generic fold. First, define a class \K{CaseFold..Aux} with a function \K{caseFold..Aux}, which perform the main work and serve for managing the two recursion cases. As before, we will use the synonyms \K{CaseFold} and \K{caseFold} to abbreviate their use with the application of \K{Equal a x}:
\begin{lstlisting}[style=fancy]
class CaseFold.Aux (eq :: Bool)
                  (a :: *) (b :: *) (x :: *) (y :: *) where
  caseFold.Aux :: Algebra a b -> I x -> I y

instance CaseFold.Aux 'False a b x x where
  caseFold.Aux _ = id
instance GFold a b => CaseFold.Aux 'True a b a b where
  caseFold.Aux f = I . applyAlgebra @a f . foldSOP . from . unI
    where
      foldSOP = trans_SOP (Proxy @(CaseFold a b)) $
                          caseFold @a @b f
\end{lstlisting}
The function \K{caseFold..Aux} is meant to operate on the elements of a product constituting a part of a datatype's representation. For constants, there are no recursive calls, so an operation for that case is trivially the identity function. Whereas in the case of the recursion point, this turns the element to its representation, then recursively processes that, and applies the given algebra to the result. This employs the function \K{applyAlgebra} that applies the algebra to the representation gained as a result of processing:
\begin{lstlisting}[style=fancy]
applyAlgebra :: ∀a b. ApplyAlgebra a b
             => Algebra a b -> SOP I (AlgCode a b) -> b
\end{lstlisting}
That representation has the code, calculated from the origin code of \K{a} by replacing its recursive occurences therein with the algebra's result type \K{b}. \K{AlgCode a b} here is a type synonym for the result of this computation. We omit the implementation of \K{applyAlgebra}, as well as of the type-level machinery of \K{AlgCode} and \K{ApplyAlgebra}, for the sake of space.

The type applications \K{@a} and \K{@b} appear in the lines above, because they are needed to calculate the \K{Algebra a b} type. Also, the constraint \K{GFold a b} on the second instance of \K{CaseFold..Aux} is necessary to use \K{applyAlgebra}, \K{caseFold}, and \K{from}:
\begin{lstlisting}[style=fancy]
type GFold a b = (Generic a
                 , AllZip2 (CaseFold a b) (Code a) (AlgCode a b)
                 , ApplyAlgebra a b)
\end{lstlisting}

The generic fold function \K{gfold} is defined using \K{caseFold}:
\begin{lstlisting}[style=fancy]
gfold :: ∀a b. GFold a b => Algebra a b -> a -> b
gfold f = unI . caseFold @a @b f . I
\end{lstlisting}

This can be conveniently used with various algebras that are now of type \K{NP}, if we define an infix operator \K{&} for constructing products:
\begin{lstlisting}[style=fancy]
infixr 1 &
x & ys = I x :* ys
\end{lstlisting}
The call of \K{gfold} with \K{example}, defined above in this subsection,
\begin{lstlisting}[style=fancy]
gfold (id & (+) & (*) & Nil) example
\end{lstlisting}
again yields 15.

@@

This machinery employs the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec..Aux}. The predicate \K{P} generalizes \K{Equal} to an arbitrary relation on types, so that one can model, e.g., subtyping for some domain-specific language. The introduction of the associated type class \K{DispatchRec} provides a handy interface for dispatch. The dispatch function \K{dispatchRec} decides on one of the functions \K{f.1} and \K{f.2}---depending on whether the predicate \K{P} is true for types \K{a.1}, \dots, \K{a.k}, \K{b}. Here, the type \K{b} is supposed to be an instance of \K{Generic}.

The type \K{X b}, associated with \K{b}, depends of the kind of the generic function. All generic functions may be classified into three categories~\cite{Rodriguez2008}: \emph{consumers}, \emph{transformers}, and \emph{producers}. Given type \K{b}, an instance of \K{Generic}, they turn \K{b} into a constant type, \K{b} into \K{b} with a changed value, and a constant type into \K{b}, respectively. For consumer functions, such as \K{gshow} (Section~\ref{subsec:gshow}), the type \K{X b} typically is instantiated with \K{b -> T} for some datatype \K{T}. For transformers, such as \K{gcompos} (Section~\ref{subsec:rec-schemes}), \K{X b} typically takes the form \K{S -> I a.i -> I a.j}. Here, \K{a.i}, \K{a.j} for $\textcolor{cbasic}{i} \neq \textcolor{cbasic}{j}$ are from \K{a.1 ... a.n}, and are supposed to be equal to \K{b} at recursion points, and \K{S} is some type. This suggests that the transformer function changes \K{b}, employing values of type \K{S}. The pattern fits equally well for defining producers, such as \K{garbitrary} from the \textsf{basic-sop} package, where \K{X b} would be instantiated with \K{Gen b}.

@@

\section{Generic generic programming}
\label{sec:ggp}
